---
title: "Lab 4 - Gr. 14 - Bioinformatics (732A93)"
author: Julius Kittler (julki092), Stefano Toffol (steto820),
        Saewon Jun (saeju204), Maximilian Pfundstein (maxpf364)
output:
  html_document:
    df_print: paged
  pdf_document: default
link-citations: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(fig.width = 7, fig.height = 3, echo = FALSE, 
                      warning = FALSE, message = FALSE) 

# All privided Links:
# R Bio: Untangling Genomes
# https://www.bioconductor.org/help/course-materials/ 2015/Uruguay2015/
# Step by Step HUVEC and OVE
# https://www.bioconductor.org/help/course-materials/2015/Uruguay2015/ day3-gene.expression.html 
# Cell Data:
# ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE20nnn/GSE20986/suppl/
# Description of Cell Data:
# https://www.ncbi.nlm.nih.gov/geo/query/acc. cgi?acc=GSE20986
# Additional
#  https://www.bioconductor.org/help/course-materials/ 2015/Uruguay2015/day5-data_analysis.html
# Explanations Graphic
# https://www.bioconductor.org/help/course-materials/2015/Uruguay2015/V6-RNASeq.html

library(ggplot2)

# Use this if BiocManager is not installed
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#library("BiocManager")

# BiocManager packages
# BiocManager::install("GEOquery", version = "3.8")
# BiocManager::install("simpleaffy", version = "3.8")
# BiocManager::install("RColorBrewer", version = "3.8")
# BiocManager::install("affyPLM", version = "3.8")
# BiocManager::install("limma", version = "3.8")
# BiocManager::install("annotate", version = "3.8")
# BiocManager::install("hgu133plus2.db", version = "3.8")
library(GEOquery)
library(simpleaffy)
library(RColorBrewer)
library(affyPLM)
library(limma)
library(hgu133plus2.db)
library(annotate)

```

Provided source code we have to explain.

```{r, echo = FALSE}

# Get the Data
#x = getGEOSuppFiles("GSE20986")
#x

# Untar and Unzip
# DONT ADD THE FILE TO GIT IT'S 60MB!
#untar("GSE20986/GSE20986_RAW.tar", exdir = "data")
#cels = list.files("data/", pattern = "[gz]")
#sapply(paste("data", cels, sep = "/"), gunzip)

# Creating phenodata
# It's a matrix with two columns. Each row has the same entries in their cells,
# which is the filename. The first column is called 'Name' and the second one
# is called 'FileName'.
phenodata = matrix(rep(list.files("data", pattern = "[CEL]"), 2), ncol =2)
class(phenodata)
phenodata <- as.data.frame(phenodata)
colnames(phenodata) <- c("Name", "FileName")
# Adding a new column with the target
phenodata$Targets <- c("iris", 
                       "retina", 
                       "retina", 
                       "iris", 
                       "retina", 
                       "iris", 
                       "choroid", 
                       "choroid", 
                       "choroid", 
                       "huvec", 
                       "huvec", 
                       "huvec")
# Writes the dataframe to a .txt file
write.table(phenodata, "data/phenodata.txt", quote = F, sep = "\t",
            row.names = F)
# Now the data is read again and a boxplot is created
celfiles <- read.affy(covdesc = "phenodata.txt", path = "data")
# Creates a boxplot for the log base 2 intensities including pm (perfect) and 
# mm (mismatch)
boxplot(celfiles)

# Create "nice looking" color palettes. So this array actually has hex-color
# values inside.
cols = brewer.pal(8, "Set1")
# Help page: exprs,AffyBatch-method
# The columns are the different .CEL files and hold the data. It's a class re-
# presentation for the probe level data. The main component are the intensities
# from multiple arrays of the save CDF type.
eset <- exprs(celfiles)
samples <- celfiles$Targets
colnames(eset)
# The colnames are set to our targets
colnames(eset) <- samples
# The .CEL files and their content is being plotted. It's basically the same
# plot as before but this time with fancy colors
boxplot(celfiles, col = cols, las = 2)
# Creating a distance matrix
distance <- dist(t(eset), method = "maximum")
# "Hierarchical cluster analysis on a set of dissimilarities and methods for analyzing it."
clusters <- hclust(distance)
# Plot the clusters in a tree diagram
# (https://en.wikipedia.org/wiki/Dendrogram)
plot(clusters)
# "Robust Multi-Array expression measure using sequence infornmation"
# Converts an AffyBatch into an ExpressionSet by using RMA (robust multi-array)
# expression measure with the help of probe sequence
celfiles.gcrma = gcrma(celfiles) # Biobase / ExpressionSet

# These are two Boxplots showing the different .CELS before and after normali-
# zation
par(mfrow=c(1,2))
boxplot(celfiles.gcrma, col = cols, las = 2, main = "Post-Normalization");
boxplot(celfiles, col = cols, las = 2, main = "Pre-Normalization")

# Turns of the "devices"
dev.off()

#Create a distance matrix
distance <- dist(t(exprs(celfiles.gcrma)), method = "maximum")
# Performs a hierachical cluster analysis.
clusters <- hclust(distance)
# Plots the clusters, which is a Dendogram again
plot(clusters)

phenodata

samples <- as.factor(samples)
# Creates a design/model matrix by expanding factors.
design <- model.matrix(~0+samples)
# "sampleschoroid" "sampleshuvec"   "samplesiris"    "samplesretina"
colnames(design)

# Rename from "sample" to this
colnames(design) <- c("choroid", "huvec", "iris", "retina")
design

# "Construct the contrast matrix corresponding to specified contrasts of a se
# of parameters."
contrast.matrix = makeContrasts(
              huvec_choroid = huvec - choroid, 
              huvec_retina = huvec - retina, 
              huvec_iris <- huvec - iris, 
              levels = design)

# Fits a linear model for each gene given a series of arrays
# Results from the RMA and the contrast.matrix
fit = lmFit(celfiles.gcrma, design)
# Use the model to fit microdata data (coefficients, errors)
huvec_fit <- contrasts.fit(fit, contrast.matrix)
# Calculates t-statistics, F-statistics and logodds for the fitted data
huvec_ebay <- eBayes(huvec_fit)

# Creates a list of probe name for 100000 entries. topTable tkaes the top-ranked
# genes from a linear fit
probenames.list <- rownames(topTable(huvec_ebay, number = 100000))
# Get symbols
getsymbols <- getSYMBOL(probenames.list, "hgu133plus2")
# And get the 100000 by the coefficient huvec_choroid
results <- topTable(huvec_ebay, number = 100000, coef = "huvec_choroid")
# Add the symbols to this
results <- cbind(results, getsymbols)

# Prints the summary of our results
summary(results)

# "Extract Data" depending on the p value and logFC
results$threshold <- "1"
a <- subset(results, adj.P.Val < 0.05 & logFC > 5)
results[rownames(a), "threshold"] <- "2"
b <- subset(results, adj.P.Val < 0.05 & logFC < -5)
results[rownames(b), "threshold"] <- "3"
table(results$threshold)

# Make a volcano plot
volcano <- ggplot(data = results, 
                  aes(x = logFC, y = -1*log10(adj.P.Val), 
                      colour = threshold, 
                      label = getsymbols))

volcano <- volcano + 
  geom_point() + 
  scale_color_manual(values = c("black", "red", "green"), 
                 labels = c("Not Significant", "Upregulated", "Downregulated"), 
                 name = "Key/Legend")

volcano + 
  geom_text(data = subset(results, logFC > 5 & -1*log10(adj.P.Val) > 5),
            aes(x = logFC, y = -1*log10(adj.P.Val), colour = threshold,
                label = getsymbols)  )

```



# Assignment 1

**Task:**

- Run all the R code and reproduce the graphics.
- Go carefully through the R code and explain in your words what each step does.

```{r, echo = FALSE}
# ------------------------------------------------------------------------------
# Question 1
# ------------------------------------------------------------------------------

```


## Assignment 2

**Task:**

- Present the variables versus each other original, log–scaled and MA–plot for each considered pair both before and after normalization.
- A cluster analysis is performed on the page but not report. Present plots and also draw heatmaps.

```{r}
# ------------------------------------------------------------------------------
# Question 2
# ------------------------------------------------------------------------------

```

# Assignment 3

**Task:**

- Provide volcano plots for the other pairs.
- Indicate significantly differentially expressed genes.
- Explain how they are found.

```{r}
# ------------------------------------------------------------------------------
# Question 3
# ------------------------------------------------------------------------------

```

## Assignment 4

**Task:**

- Try to find more information on the genes that are reported to be significantly differentially expressed. Report in your own words on what you find.
- Report all the Gene Ontology (GO) terms associated with each gene.
  - Are any of the GO terms common between genes?
  - If so do the common GO terms seem to be related to anything particular?
- Try to present GO analysis in an informative manner, if possible visualize.


```{r}
# ------------------------------------------------------------------------------
# Question 4
# ------------------------------------------------------------------------------

```


# Appendix

```{r, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}

```