---
title: "Lab 4 - Gr. 14 - Bioinformatics (732A93)"
author: Julius Kittler (julki092), Stefano Toffol (steto820),
        Saewon Jun (saeju204), Maximilian Pfundstein (maxpf364)
output:
  pdf_document: default
  html_document:
    df_print: paged
link-citations: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(fig.width = 7, fig.height = 3, echo = FALSE, 
                      warning = FALSE, message = FALSE) 

# All privided Links:
# R Bio: Untangling Genomes
# https://www.bioconductor.org/help/course-materials/ 2015/Uruguay2015/
# Step by Step HUVEC and OVE
# https://www.bioconductor.org/help/course-materials/2015/Uruguay2015/ day3-gene.expression.html 
# Cell Data:
# ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE20nnn/GSE20986/suppl/
# Description of Cell Data:
# https://www.ncbi.nlm.nih.gov/geo/query/acc. cgi?acc=GSE20986
# Additional
#  https://www.bioconductor.org/help/course-materials/ 2015/Uruguay2015/day5-data_analysis.html
# Explanaitions Graphic
# https://www.bioconductor.org/help/course-materia 2015/Uruguay2015/V6-RNASeq.html

```

# Assignment 1

**Task:**

- Run all the R code and reproduce the graphics.
- Go carefully through the R code and explain in your words what each step does.

```{r, echo = FALSE}
# ------------------------------------------------------------------------------
# Question 1
# ------------------------------------------------------------------------------

```


## Assignment 2

**Task:**

- Present the variables versus each other original, log–scaled and MA–plot for each considered pair both before and after normalization.
- A cluster analysis is performed on the page but not report. Present plots and also draw heatmaps.

```{r}
# ------------------------------------------------------------------------------
# Question 2
# ------------------------------------------------------------------------------

```

# Assignment 3

**Task:**

- Provide volcano plots for the other pairs.
- Indicate significantly differentially expressed genes.
- Explain how they are found.

```{r}
# ------------------------------------------------------------------------------
# Question 3
# ------------------------------------------------------------------------------

```

## Assignment 4

**Task:**

- Try to find more information on the genes that are reported to be significantly differentially expressed. Report in your own words on what you find.
- Report all the Gene Ontology (GO) terms associated with each gene.
  - Are any of the GO terms common between genes?
  - If so do the common GO terms seem to be related to anything particular?
- Try to present GO analysis in an informative manner, if possible visualize.


```{r}
# ------------------------------------------------------------------------------
# Question 4
# ------------------------------------------------------------------------------

```


# Appendix

```{r, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}

```