---
title: "Lab 1 - Gr. 14 - Bioinformatics (732A93)"
output: pdf_document
link-citations: true
author: tbd
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(fig.width = 7, fig.height = 3, echo = FALSE) 

library(dplyr)
library(tidyr)
library(magrittr)

```

# Assignment 1

## 1.1

Some text and equation here: 

\begin{equation}
  S = \sqrt{\frac{\sum_{r,s}(d_{rs} - \hat{d}_{rs})^2}{\sum_{r,s} d_{rs}^2}}.
\end{equation}


```{r}
# ------------------------------------------------------------------------------
# Question 1.1
# ------------------------------------------------------------------------------

```

## 1.2

```{r}
# ------------------------------------------------------------------------------
# Question 1.2
# ------------------------------------------------------------------------------

```

# Assignment 2

Note that, by convention, a *coding strand* is used when displaying a DNA sequence.
"A coding strand, is the segment within double-stranded DNA that runs from 5' to 3', and which is complementary to the antisense strand of DNA, or template strand, which runs from 3' to 5'" (https://en.wikipedia.org/wiki/Sense_strand).

https://www.ncbi.nlm.nih.gov/nuccore/CU329670


## 2.1

```{r}
# ------------------------------------------------------------------------------
# Question 2.1
# ------------------------------------------------------------------------------

```

## 2.2

```{r}
# ------------------------------------------------------------------------------
# Question 2.2
# ------------------------------------------------------------------------------

```

# Assignment 3

## 3.1

```{r}
# ------------------------------------------------------------------------------
# Question 3.1
# ------------------------------------------------------------------------------

```

## 3.2

```{r}
# ------------------------------------------------------------------------------
# Question 3.2
# ------------------------------------------------------------------------------

```

# Appendix

```{r, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}

```